Investigation of ant cuticle dataset using image texture analysis methods

### 2022-07-22 paper/main.tex
pre submission

Because ants lack bold coloration and complex wing venation patterns used to
identify other insect species, microsculpturing has taken a prominent role in
ant identification. To aid in species identification, ant taxonomists have
developed nearly 100 terms to describe  fine differences in microsculpturing
patterns \cite{harris_glossary_1979}. Despite the prominent role of
microsculpturing in ant identification, however, almost nothing is known about
how these patterns evolved or their function. Similar microsculpturing patterns
in other organisms have been found to reduce abrasion, increase stiffness, and
perform as structural antibiotics \cite{zhiwu_erosion_2012, han_efficient_2015,
    rees_form_1975, chung_impact_2007,hasan_selective_2013}. Developing a better
understanding of the function of microsculpturing in ants could provide insight
into their evolution and biology as well as lead to the development of
bio-inspired technologies \cite{penick_comparative_2022}.

One challenge for functional investigations of cuticle microsculpturing is the
complexity of terms used to describe these patterns. Complex terminology may be
useful for describing fine differences between closely related species, but
broader comparisons require a simplified classification scheme. We recently
developed a classification system for microsculpturing that consists of only
five categories: smooth, striate, punctate, reticulate, and tuberous. Using
these categories to classify 11,722 ant species, we found that the earliest ants
were likely smooth and that cuticle microsculpturing independently evolved later
in multiple lineages. An even simpler classification scheme is to divide ants
into only two categories: smooth and rough. Because microsculptuing has evolved
numerous times in ants, this binary classification scheme is sufficient to
detect conditions that select for any sculpturing at all compared to a smooth
state.

@@

For the feasibility study in this ant image datasets, we use both K-means and
K-views in our experiments as both are statistical clustering methods. The
experiments may indicate how well the statistical approach will do in this type
of textural images. In pre-processing the dataset, we apply the Gabor filter to
transform the gray-scale ant image into four Gabor-filtered channels, subtract
the gray-scale image from each channel, take the absolute value of the resulting
images, and then normalize them. This pre-processing will highlight some of
characteristics features such as edge-like in the image. This step is very
useful in feeding the datasets to the K-views algorithms. A 10-channel Gabor
filter is applied in the pre-processing step for the K-means algorithm. The
theta $\theta$ parameters in the Gabor filters used are multiples of $\pi/10$
radians starting from $0$ to $\pi$ radians. The Gabor filter lambda $\lambda$
parameter is equal to $\pi$.

>>>>

Because ants lack bold coloration and complex wing venation patterns used to
identify other insect species, microsculpturing has taken a prominent role in
ant identification. To aid in species identification, ant taxonomists have
developed nearly 100 terms to describe  fine differences in microsculpturing
patterns \cite{harris_glossary_1979, bolton_identification_1994}. Despite the
prominent role of microsculpturing in ant identification, however, almost
nothing is known about how these patterns evolved or their function. Similar
microsculpturing patterns in other organisms have been found to reduce abrasion,
increase stiffness, and perform as structural antibiotics
\cite{zhiwu_erosion_2012, han_efficient_2015, rees_form_1975,
    chung_impact_2007,hasan_selective_2013}. Developing a better understanding of
the function of microsculpturing in ants could provide insight into their
evolution and biology as well as lead to the development of bio-inspired
technologies \cite{penick_comparative_2022}.

One challenge for functional investigations of cuticle microsculpturing is the
complexity of terms used to describe these patterns. Complex terminology may be
useful for describing fine differences between closely related species, but
broader comparisons require a simplified classification scheme. We recently
developed a classification system for microsculpturing that consists of only
five categories: smooth, striate, punctate, reticulate, and tuberous
\cite{hellenbrand_ant_2022}. Using these categories to classify 11,722 ant
species, we found that the earliest ants were likely smooth and that cuticle
microsculpturing independently evolved later in multiple lineages
\cite{hellenbrand_ant_2022}. An even simpler classification scheme is to divide
ants into only two categories: smooth and rough. Because microsculptuing has
evolved numerous times in ants, this binary classification scheme is sufficient
to detect conditions that select for any sculpturing at all compared to a smooth
state.

@@

For the feasibility study in this ant image datasets, we use both K-means and
K-views in our experiments as both are statistical clustering methods. The
experiments may indicate how well the statistical approach will do in this type
of textural images. In pre-processing the dataset, we apply the Gabor filter to
transform the gray-scale ant image into four Gabor-filtered channels, subtract
the gray-scale image from each channel, take the absolute value of the resulting
images, and then normalize them. This pre-processing will highlight some of
characteristics features such as edge-like in the image. This step is very
useful in feeding the datasets to the K-views algorithms. A 4-channel Gabor
filter is applied in the pre-processing step for the K-means algorithm. The
theta $\theta$ parameters in the Gabor filters used are multiples of $\pi/4$
radians starting from $0$ to $\pi$ radians. The Gabor filter lambda $\lambda$
parameter is equal to $\pi$.
